---
<<<<<<< HEAD
title: "AGE AND SMOKING: A COMPREHENSIVE ANALYSIS USING THE FREQUENTIST AND BAYESIAN METHOD "
output: html_document
date: "2024-10-5"
AUTHORS: Kiana Philpott, Chizurum Ewelike, LAYO Faloseyi
=======
title: "STATISTICS 3910 Project"
author: "Kiana, Chizurum, Layo"
date: "Last updated on `r Sys.Date()`"
output:
html_document:
fig_caption: yes
theme: lumen
toc: yes
toc_depth: 2
df_print: kable
toc_float:
collapsed: no
>>>>>>> ba637b0 (we did the Bayesian and Frequentist analysis)
---

```{r, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# potential packages of interest
packages = c("ALSM", "mosaic", "car", "leaps", "tidyverse", "here",
"BayesFactor")


# load all packages here:
library(readr)
library(dplyr)
library(ggplot2)
library(janitor)
library(BayesFactor)

# set seed value of random number generator to get "replicable" random numbers (in case?).
set.seed(100)
```

<<<<<<< HEAD
## R Markdown
**Research Hypothesis**:
 <br> 
  This study aims to analyze the age demographics of those who smoke. The null hypothesis, H0, assumes that the average age of smokers is 60 years or younger, while the alternative hypothesis, H1, assumes that the average age of individuals who smoke is greater than 60 years.

**Background**:
 <br> 
  The nature of smoking remains a public health issue, and seems to affect people across varying age groups. This research aims to determine whether older adults are the most likely to smoke. In doing so, we might be able to recorgnize patterns that could inform the public health sector and contribute to reduced smoking rates. This study is particularly interesting as it highlights how societal attitudes towards smoking could evolve over time 

<<<<<<< HEAD
**Data Sketch**:
 <br> 
  The data set includes survey data on smoking habits from the United Kingdom. The data set can be used for analyzing demographic characteristics of smokers. 

**Size**: 
 <br> 
There are 12 variables and there are 1691 observations
=======
Data Sketch:
## kiana trial

Size: second trial
>>>>>>> 92d6b980451b413a13d97482f11a7f3450c02a64

**Source**: https://www.kaggle.com/datasets/utkarshx27/smoking-dataset-from-uk?resource=download

**Analysis**: 
 <br> 
  The tentative analysis plan is as follows:
  <br> 
  <br> 
     | **1**. Check for the need of data cleaning. 
      <br> 
      <br> 
     | **2**. Get familiar with the data set itself. 
      <br> 
      <br> 
     | **3**. Create visualizations to show the distributions to show the distribution of age and smoking to enhance understanding
      <br> 
      <br> 
     | **4**. Frequentist approach:
      <br> 
      <br> 
          - Conduct a T-Test that compares the mean age of smokers to the age variable 
           <br> 
           <br> 
          - Validate assumptions for the T-Test (normality & homogeneity of variances)
           <br> 
           <br> 
          - Calculate a 95% confidence interval for the mean age of smokers to further assess the accuracy of our estimate
           <br> 
           <br> 
      | **5**. Bayesian approach:
       <br> 
       <br> 
          - Conduct a Bayestian T-Test using prior distribution 
           <br> 
           <br> 
          - Analyze the posterior distribution to estimate the mean age & credible interval to help assess the probability that the               mean age of smokers is greater than 60 years 
           <br> 
           <br> 
          - Conduct a prior sensitivity analysis by testing different priors to understand the impact they have on the results
           <br> 
           <br> 
      | **6**. Compare the results from each approach (includes: interpretations & visualizations)
       <br> 
       <br> 
      | **7**. Document the findings & results
       <br> 
      <br> 
      
      
      
      
      
      

=======
# Big-picture

## Research question

This study aims to analyze the age demographics of those who smoke. The null hypothesis, H0, assumes that the average age of smokers is 60 years or younger, while the alternative hypothesis, H1, assumes that the average age of individuals who smoke is greater than 60 years.

## Description of data
The original source: https://www.kaggle.com/datasets/utkarshx27/smoking-dataset-from-uk?resource=download
The data set includes survey data on smoking habits from the United Kingdom. The data set can be used for analyzing demographic characteristics of smokers.

```{r, message=FALSE}
# read in the csv file
smoking <- read.csv("https://drive.google.com/uc?export=download&id=1_w27s7YcT4cT6PcSTe9UA4Ul5ZASH-tt")
```



```{r}
# remove unneccesary columns from dataset
newdataset <- smoking %>% select(-amt_weekdays, -amt_weekends, -marital_status,-highest_qualification, -nationality, -gross_income, -region, -ethnicity) 

# view updated dataset
glimpse(newdataset) 
```

```{r}
# Frequentist Analysis

# check if age approximately follows a normal distribution by creating a histogram

# load necessary library
library(ggplot2)

# create a histogram of the age variable
ggplot(newdataset, aes(x = age)) +
geom_histogram(bins = 30, fill = "blue", color = "black", alpha = 0.7) +
theme_minimal() +
labs(
title = "Histogram of Smokers' Ages",
x = "Age",
y = "Frequency"
) +
geom_density(aes(y = ..density..), color = "red", linewidth = 1.2) # adds a density curve

```


```{r}
# check for outliers


# create a boxplot for the 'age' variable
boxplot(newdataset$age, 
main = "Boxplot for Age", 
outpch = 19,# change the point character for outliers
outcol = "red",# make the outliers red
outcex = 1.5,# make the outlier points larger
ylab = "Age", 
col = "lightblue", 
border = "darkblue")

# Based on our boxplot the median age appears to be between 40 and 45. the spread of the box indicates that most data is between 30 and 60 years of age.

```





```{r}
qqnorm(newdataset$age, main = "QQ-Plot of Age")
qqline(newdataset$age, col = "red", lwd = 2)

# The red line represents the expected pattern if our data was normally distributed. Our points deviate from the red line at both tail ends, indicating non-normality. The left tail shows points above the line, which suggests less extreme lower values and the right tail shows points below the line which suggests less extreme higher values. So we need to keep in mind there is a bit of deviation from normaility in our analysis. 
```





```{r}

# perform a one-sample t-test to check if the average age of smokers is 60
smokers_data <- newdataset[newdataset$smoke == "Yes", ]

# run the t-test
t_test_result <- t.test(smoking$age, mu = 60)

# extract the t-statistic and p-value
t_stat <- t_test_result$statistic
p_value <- t_test_result$p.value

# print the results
cat("t-statistic:", t_stat, "\n")
cat("p-value:", p_value, "\n")


# we get a T-Statistic of -22.30651 which suggests strong evidence against the null hypothesis
# our P-Value of nearly 0 indicates strong evidence against the null hypothesis
# we can reject the null hypothesis and conclude that the average age of smokers is significantly different from 60 (lower than 60).
```


```{r}
# Bayesian Analysis


## Load the BayesFactor package
library(BayesFactor)

# Perform the Bayesian t-test
bayesian_t_test <- ttestBF(
x = smokers_data$age,# data for smokers
mu = 60# mean under null hypothesis
)

# Extract the Bayes Factor
bf_value <- exp(bayesian_t_test@bayesFactor$bf)# Extract and exponentiate the log Bayes Factor

# Display the Bayes Factor
cat("Bayes Factor (BF10):", bf_value, "\n")

# Our Bayes Factor value of 8.901085e+67 shows evidence in favor of our alternative hypothesis. The data is extremely unlikely under our null hypothesis. Therefore, we see that the average age of smokers is not 60 or younger which strongly supports the alternative hypothesis.
```

# Analysis and Results


```{r}
# Both the Frequentest analysis and the Bayesian analysis

# Load library
library(BayesFactor)

# Ensure 'smoke' is a factor with exactly two levels ("Yes" and "No")
newdataset$smoke <- droplevels(as.factor(newdataset$smoke))

# Verify that 'smoke' has only two levels
levels(newdataset$smoke)

# Perform the Bayesian t-test comparing age between smokers and non-smokers
bayestesting <- ttestBF(
# Formula specifying age by smoke status
formula = age ~ smoke,
# Data set to analyze
data = newdataset, 
# Test for the full range, not directional
nullInterval = NULL
)

# Display the Bayes Factor result
print(bayestesting)


```


# Conclusion

# Discussion and Biases
>>>>>>> ba637b0 (we did the Bayesian and Frequentist analysis)
